---
title: "Unraveling Tornado Trends"
subtitle: "Proposal"
format: html
editor: visual
---

```{r load-pkgs,warning=FALSE,message=FALSE, include=FALSE}
library(tidyverse)
chooseCRANmirror(ind=1)
install.packages("kableExtra")
library(kableExtra)

```

## Dataset

```{r dataset, warning=FALSE,message=FALSE, include=FALSE}
tornados <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-16/tornados.csv')
```

About the data-set:

The data-set contains information about tornadoes in the United States. Each row represents a specific tornado event, and the columns provide various attributes related to each tornado event like time of occurrence, magnitude on F scale, number of injuries, fatalities, estimated property loss in dollars, longitudes and latitudes. The data-set has a total of 68693 events recorded as rows, each having 27 attributes as columns.The presented data-set was generated from a data-set created by the Storm Prediction Center of NOAA. The period of the data ranges from 1950 to 2022.

Reason for Choosing:

This data-set could be valuable for studying and analyzing tornado events in the United States. It provides a wide range of information about each tornado, including location, magnitude, impact, and other relevant details. It could be used for various analytical purposes, such as studying trends in tornado occurrences, understanding their impacts, and potentially even for building predictive models related to severe weather events.

## Questions

1\) Are there any patterns in the magnitude of tornadoes based on location or time of year? (Intensity (magnitude) of tornadoes in each state. )

2\) How many tornadoes affected multiple states, and which states are most commonly affected together?

## Analysis plan

**Question 1**: Are there any patterns in the magnitude of tornadoes based on location or time of year?

Analysis Plan:

Variables Involved:

\- "mag": \[integer\] Magnitude of tornadoes (F scale or EF scale).

\- "st": \[character\] Two-letter postal abbreviation for the states

\- "yr": \[integer\] Year (range: 1950 - 2022).

We first group the data state-wise using "st" variable and year-wise using "yr" variable, then calculate the summary statistics on both the grouped data separately. From the obtained results we check for any particular states having undergone multiple variations over time and select a few such states along with few other states for further plotting. Then for each of these states we plan to show the variations in tornado intensities over the course of time using alley maps visualization.

**Question 2:** What are the trends in the frequency of tornado-related fatalities over the years for different states in the USA?

Analysis Plan:

Variables Involved:

\- "fat": \[integer\] Number of fatalities.

\- "yr": \[integer\] Year of the tornado occurred.

\- "st": \[character\] Two-letter postal abbreviation for the state.

To analyze trends in fatalities caused by tornadoes across different states, we will focus on the variables "fat" (number of fatalities), "yr" (year), and "st" (state). First, we will aggregate the fatality data, summarizing the total fatalities for each year. Next, we will use an line/area plot to see in which years more deaths occurred. We will select those years and plot an horizontal bar plot for each state showing the number of deaths in those particular years. This approach will provide valuable insights into the trends of tornado-related fatalities in key states, aiding in disaster response planning and resources allocation.
